---
title: "EEG Utils"
author: "Matt Craddock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EEG_Utils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("EVAL_EEG_VIGNETTE"))
)
```

`eegUtils` is a package for performing basic EEG preprocessing and plotting of EEG data. Many of these functions are wrappers around existing R functions to make them behave in consistent ways and produce output that is more amenable to many of the subsequent steps in EEG analysis.

The package implements custom objects to contain EEG data and associated metadata. Some of its functions depend on data being stored in this format, but part of the philosophy of the package is that any object stored in the custom `eeg_data` and associated formats will always be convertible to a standard data.frame or tibble for subsequent use in whatever way the user desires. Plotting functions will typically work on both `eeg_data` objects and standard formats, while more complex processing functions will require an `eeg_data` or related custom object (such as `eeg_stats`).

## Basic EEG processing

There is currently suport for loading raw data in the .BDF (typically BioSemi) and .CNT (32-bit; associated with Neuroscan) file formats using the `import_raw()` command. Loading data in these formats in results in an `eeg_data` object - a structure that contains the raw data and a variety of metadata.

```{r basic-proc}
library(eegUtils)
eeg_example <- import_raw("F:\\Dropbox\\EEGData\\RawEEGData\\BasicCat\\S2B1.bdf")
eeg_example
```

The data itself is stored in the `signals` data frame within the object; the timing of events is stored in the `events` data frame, and timings for each sample are stored in the `timings` data frame. Times are stored in samples and seconds.

A common first step would be to rereference the data, which can be done using the `reref_eeg()` command. By default, if no electrodes are specified, the data will be referenced to a common average. The median rather than the mean can be used by specifiying `robust == TRUE`. Electrodes can also be excluded from the average reference calculation; they will nevertheless still be referenced to the same reference as the other channels.

```{r reref-dat}
eeg_example <- reref_eeg(eeg_example, ref_chans = "average")
eeg_example <- reref_eeg(eeg_example, ref_chans = "average", robust = TRUE)
eeg_example <- reref_eeg(eeg_example, ref_chans = "average",
                      exclude = c("EXG7", "EXG8"), robust = TRUE)
```

EEGLAB can be loaded using the `import_set()` command. The command detects whether the data is epoched or not and, by default, arranges the data in the `eeg_data` format.

Filtering can be performed using the `iir_filt()` command. This use a Butterworth filter to perform low-, high-, or band-pass or band-stop filtering of user-specified frequencies. Filtering is run twice (once forwards, once in reverse) to achieve zero-phase. (FIR filtering is currently in development, but could be performed using the `signal` package).

```{r filt-butter}
eeg_example <- iir_filt(eeg_example,
                        low_freq = .1,
                        high_freq = 40,
                        filter_order = 3) #specify a bandpass filter
```

## Creating epochs

Data can be epoched around events/triggers using `epoch_data()`, which outputs an `eeg_epochs` object. A list of the event triggers found in the data can be retrieved using `list_events()`. In this case we’ll epoch around event code 201.

```{r epoch-eeg}
list_events(eeg_example)
eeg_example <- epoch_data(eeg_example,
                          events = 201,
                          time_lim = c(-.3, .75))
```

## Plotting 

`eeg_epochs` can then be plotted using `plot_butterfly()` or `plot_timecourse()`. Both `plot_butterfly()` and `plot_timecourse()` average over epochs. `plot_timecourse()` will also average over electrodes - all electrodes if none are specified, or over any specified electrodes. Baseline correction can also be applied at this point if desired. Baseline correction can also be applied using `rm_baseline()`, or for plotting only using the `baseline` parameter in the plotting call..

```{r do-plots}
plot_butterfly(eeg_example,
               time_lim = c(-.2, .5),
               legend = FALSE)
plot_butterfly(eeg_example,
               time_lim = c(-.2, .5),
               legend = FALSE,
               baseline = c(-.1, 0))
plot_timecourse(eeg_example,
                time_lim = c(-.2, .5),
                baseline = c(-.1, 0),
                electrode = "A30") # Plot POz
plot_timecourse(eeg_example,
                time_lim = c(-.2, .5),
                baseline = c(-.1, 0),
                electrode = "B6") # Plot Fz
```

Standard channel locations can be added using the `electrode_locations()` command. Currently this only has locations for a limited number of electrodes (64 electrodes from the 10-20 system). Different built-in montages will be added, and can be specified using the `montage` parameter. In this case, the data was recorded using Biosemi with alpha-numeric channel labels (e.g. A1 rather than Fp1), which are built-in with `biosemi64alpha`. `topoplot()` can then be used to plot topographical representation of selected data. Note that it is not compulsory to use locations from electrode_locations(); if the data has x and y columns when it is a data frame, or added to `chan_info` element of the `eeg_data`/`eeg_epochs` object, then those will be used.

```{r do-topo}
eeg_example <- electrode_locations(eeg_example,
                                   montage = "biosemi64alpha")
topoplot(eeg_example,
         time_lim = c(.1, .2))
```

Note that one electrode (Fp1/A1) was malunctioning during the the recording. We can interpolate that using spherical spline interpolation (`interp_elecs()`).

```{r inter}
eeg_example <- interp_elecs(eeg_example,
                            bad_elecs = "A1")
topoplot(eeg_example,
         time_lim = c(.1, .2))
```

At any point, `eegUtils` objects can be transformed into data frames for use with functions that don't natively support them.

```{r}
library(ggplot2)
library(dplyr)
eeg_example %>%
  select_epochs(epoch_no = 1:5) %>%
  select_elecs(c("A29", "B6")) %>%
  as.data.frame(long = TRUE) %>%
  ggplot(aes(x = time, y = amplitude)) +
  geom_line(aes(group = epoch), alpha = 0.2) + 
  stat_summary(fun.y = mean,
               geom = "line",
               size = 2,
               aes(colour = electrode)) + 
  facet_wrap(~electrode) + 
  theme_classic()
```

# Tidyverse functions

In addition, there are overloaded versions of some `dplyr` functions that operate on the `signals` element of `eeg_data` and `eeg_epochs` objects. For example, `select()` can be used to choose particular electrodes, and `filter()` can be used to filter out epochs or timepoints. `mutate()` can be used to add new columns (e.g. creating ROIs from multiple electrodes).

```{r}

eeg_example %>%
  mutate(occipital = (A29 + A27 + B32) / 3) %>%
  select(A29, B6, occipital) %>%
  filter(epoch <= 60, time < .5) %>%
  as.data.frame(long = TRUE) %>%
  ggplot(aes(x = time, y = amplitude)) +
  geom_line(aes(group = epoch), alpha = 0.2) + 
  stat_summary(fun.y = mean,
               geom = "line",
               size = 2,
               aes(colour = electrode)) +
  facet_wrap(~electrode) + 
  scale_colour_viridis_d() +
  theme_classic()
```
