---
title: "Introduction"
author: "Matt Cradock and Eryk Walczak"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This package was created to allow the analysis of EEG data in R. The code in this vignette was previously published on Matt's [blog](http://www.mattcraddock.com/).

## Downloading a bdf

Two example Biosemi bdf files can be downloaded from the producer's [website](https://www.biosemi.com/faq/file_format.htm).
To download and unzip example bdf files use:

```{r}
url_bdf_zip <- "https://www.biosemi.com/download/BDFtestfiles.zip"
temp <- tempfile()
download.file(url_bdf_zip, temp)
unzip(temp)
unlink(temp)
```

## Loading a bdf file

```{r, warning = FALSE, message = FALSE}
library(eegUtils)
library(dplyr)

bdf_filename <- "Newtest17-256.bdf"
bdf_data <- import_raw(bdf_filename)
```

The file is now loaded. You can see its structure

```{r}
str(bdf_data)
```

## Pre-processing

eegUtils allows loading the bdf files, selecting channels

```{r}
my_data <- bdf_data %>%
  select_elecs(c("A1", "A2"), keep = FALSE) %>% # remove channels
  reref_eeg(ref_chans = "average") %>% # re-reference
  iir_filt(low_freq = 0.1, high_freq = 40) %>% # band-pass filter
  epoch_data(255, c(-.2,.2)) # epoch: choose a trigger and time limits
```

## Figures

Create a butterfly plot showing data from all the electrodes 

```{r}
plot_butterfly(my_data, baseline = c(-.2, 0), time_lim = c(-.1, .2), legend = FALSE)
```

Or the timecourse of a single electrode

```{r}
plot_timecourse(my_data, electrode = "A3", baseline = c(-.2,0), time_lim = c(-.1, .2))
```

How about an EEGLAB style ERP image?

```{r}
erp_image(my_data, electrode = "A9")
```
