---
title: "Linear Modelling"
author: "Matt Craddock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = nzchar(Sys.getenv("EVAL_EEG_VIGNETTE"))
)
```


```{r}
library(eegUtils)
library(R.matlab)
library(ggplot2)
library(tidyr)
```

Here we show how to fit a linear model to a single subject's data. We'll use the data from the LIMO EEG dataset. [S1](https://datashare.is.ed.ac.uk/handle/10283/2189?show=full)



```{r}
limo_test <- import_set("limo_dataset_S1.set")
limo_cont <- R.matlab::readMat("continuous_variable.mat")
limo_cat <- readr::read_csv("categorical_variable.txt",
                            col_names = c("cond_lab"))
```

The linear model fitting function in `eegUtils` uses the `epochs` field of the data structures. To use these for linear modelling here, we'll add the categorical and continuous predictors to the `epochs` structure.

```{r}
epochs(limo_test) <- dplyr::mutate(epochs(limo_test),
                            phase_coherence = unlist(limo_cont),
                            face = factor(limo_cat$cond_lab,
                                          levels = c(1, 2),
                                          labels = c("Face_A",
                                                     "Face_B")))
epochs(limo_test)

```

The first argument to `fit_glm()` is a standard R formula, using Wilkinson notation. Unusually, the left hand side of the `~` is omitted, since we'll be fitting the model to every electrode at every timepoint. On the right hand side, we put our predictors. In this case, we have two: `face`, a categorical predictor; and `phase_coherence`, a continuous predictor. The appropriate formula is thus `~face + phase_coherence`. 

The function fits whatever contrasts. R's default contrasts are *treatment* or *dummy* contrasts, so we' expect the model to return three coefficients: an *intercept*, which will be the amplitude when the categorical predictor is at its first level (i.e. Face A); a *face* term, which will be the *difference* in amplitude from the *intercept* when the level of Face is `Face B`; and a *phase_coherence* term, which will be the increase in amplitude when image phase coherence increases from 0 to 1.

```{r}
fitted_model <- fit_glm(~ face + phase_coherence,
                        data = limo_test)
as.data.frame(fitted_model,
              long = TRUE) %>%
  ggplot(aes(x = time,
             y = amplitude,
             colour = electrode)) +
  geom_line() +
  facet_wrap(~coefficient,
             scales = "free") +
  theme(legend.position = "none")
```

An alternative paramterization would be to remove the intercept term. The function would then return separate coefficients represent `Face A` and `Face B`. This would be particularly helpful if you are planning to take these coefficients forwards to a second-level analysis.

```{r}
fitted_model_no_int <- fit_glm(~0 + face + phase_coherence,
                        data = limo_test)
as.data.frame(fitted_model_no_int,
              long = TRUE) %>%
  ggplot(aes(x = time,
             y = amplitude,
             colour = electrode)) +
  geom_line() +
  facet_wrap(~coefficient,
             scales = "free") +
  theme(legend.position = "none")
```

Continuous predictors can be rescaled using the `scale()` function, which converts them to z-scores (i.e. standard deviation units).

```{r}
fitted_model_zscore <- fit_glm(~0 + face + scale(phase_coherence),
                        data = limo_test)
as.data.frame(fitted_model_zscore,
              long = TRUE) %>%
  ggplot(aes(x = time,
             y = amplitude,
             colour = electrode)) +
  geom_line() +
  facet_wrap(~coefficient,
             scales = "free") +
  theme(legend.position = "none")
```

The function also provide additional information. For example, we can get the $r^2$ value representing model fit:

```{r}
fitted_model$r_sq %>%
  pivot_longer(cols = channel_names(limo_test),
               names_to = "electrode",
               values_to = "r_sq") %>%
  ggplot(aes(x = time,
             y = r_sq,
             colour = electrode)) +
  geom_line() +
  theme(legend.position = "none")
```
