---
title: "Building a topographical plot"
author: "Matt Craddock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a topographical plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(eegUtils)
```

`eegUtils` has convenience functions for plotting and styling topographical scalp maps, which are very frequently used when exploring EEG and MEG data.

The `topoplot()` function uses `ggplot2` to create a such a topographical map, with a cartoon head, ears and nose, and points or text labels indicating electrode positions or names. 

Here's an example using the built-in `demo_epochs` dataset.

```{r topo-orig, fig.height = 4, fig.width = 4}
topoplot(demo_epochs)
```

`topoplot()` has many different parameters that can be used to alter its appearance and select different parts of the data to be plotted. For example, here we select a specific time period to plot (the function averages over points within these limits), change the plotting so that all electrodes and interpolated surfaces are within the cartoon head, and highlight a specific electrode in white.

```{r fig.height = 4, fig.width = 4}
topoplot(demo_epochs,
         time_lim = c(.1, .2),
         palette = "viridis",
         highlights = c("B5"),
         interp_limit = "head")
```

`topoplot()` operates on several different types of `eegUtils` object, and has different default settings and behaviour for those objects. For example, for ICA decompositions, it plots the mixing matrix for each component.

```{r get-ica, message = FALSE, fig.height = 4, fig.width = 4}
demo_ica <- run_ICA(demo_epochs, pca = 10)
topoplot(demo_ica, component = 2)
```

## ggplot2 extensions - geom_topo()

For those who desire more control over the plots and are familiar with `ggplot2`, there are several custom geoms to help you do this. As a starter, the `geom_topo()` creates a full topographical plot, with cartoon head and points indicating coordinate locations. This is based on `ggplot2`'s `geom_raster()`, and has such has many parameters in common. But unlike `geom_raster()` it can handle an irregular grid.

`fortify` methods are included in the package so that most `eegUtils` objects can be passed directly to `ggplot()`. For more complete control, one might wish to convert the objects using `as.data.frame()` first, so that any desired manipulations such as averaging can be conducted first. 

```{r default-geom, fig.height = 4, fig.width = 4}
library(ggplot2)
ggplot(demo_epochs,
       aes(x = x,
           y = y,
           fill = amplitude)) +
  geom_topo()
```

```{r fig.height = 4, fig.width = 4}
ggplot(demo_epochs,
       aes(x = x,
           y = y,
           fill = amplitude)) +
  geom_topo() + 
  scale_fill_distiller(palette = "RdBu") + 
  theme_void() + 
  coord_equal() + 
  labs(fill = expression(paste("Amplitude (", mu,"V)")))
```

There are several different parameters that can be passed to `geom_topo()` to change the appearance of the plot.

```{r fig.height = 4, fig.width = 4}
ggplot(demo_epochs,
       aes(x = x,
           y = y,
           fill = amplitude,
           label = electrode)) +
  geom_topo(grid_res = 200,
            interp_limit = "head",
            chan_markers = "text",
            chan_size = 4) + 
  scale_fill_distiller(palette = "RdBu") + 
  theme_void() + 
  coord_equal() + 
  labs(fill = expression(paste("Amplitude (", mu,"V)"))) 
```

## Building from the ground up

For those who want *even more* control, individual constituent parts can be added to the plots as below.

`stat_scalpmap()` adds an interpolated surface from an irregular grid of electrodes, rounded off as a circular shape. `geom_head()` adds a cartoon headshape. `geom_channels()` is a wrapper round `geom_text()` and `geom_point()` that prevents overplotting when submitting an `eegUtils` object directly, and can be used to plot channel locations as points or text labels. `geom_mask()` can be used to draw a circular masking ring around the edge of the interpolated surface, to smooth any rough edges. This is useful when using a low grid resolution, which is advisable when using facetting or animation. All of these `geoms` are customisable as standard `ggplot2` geoms are.

```{r step-by-step}
ggplot(demo_epochs, 
       aes(x = x, 
           y = y,
           fill = amplitude)) + 
  stat_scalpmap(grid_res = 60) +  #based on geom_raster()
  geom_mask(scale_fac = 1.4) +    #based on geom_path()
  geom_head() +                   #based on geom_path() and geom_curve()
  geom_channels() +               #based on geom_point() and geom_text()
  scale_fill_viridis_c(limits = c(-6, 6), oob = scales::squish) +
  coord_equal()
```
